Hierarchical Principal Manifold Estimation

### 2025-03-18 manuscript/main.tex
rename paper directory

A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations. To develop an estimate of the manifold underlying the hippocampi of participants in the Alzheimer's Disease Neuroimaging Initiative (ADNI) study, it is necessary to recognize the basic nesting structure of the data, in which brain images are nested within participants belong to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments. These subgroups are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population. In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when estimating the manifolds underlying such data. We further propose an approach for estimation of the underlying manifold at each level of the hierarchy, allowing for statistical testing of differences between those low-dimensional manifolds.

>>>>

A method to accurately and efficiently estimate the manifolds underlying data at the group level must take advantage of the grouped structure of observations. To develop an estimate of the manifold underlying the hippocampi of participants in the Alzheimer's Disease Neuroimaging Initiative (ADNI) study, it is necessary to recognize the basic nesting structure of the data, in which brain images are possibly nested within participants belong to a latent disease subgroup where patients tend to have more related symptom experiences or underlying physiological developments. These subgroups are in turn nested within study groups based on Alzheimer's disease status, which are finally nested within the greater study population. In this paper, we develop a manifold learning approach that appropriately accounts for this hierarchical structure when estimating the manifolds underlying such data. We further propose an approach for estimation of the underlying manifold at each level of the hierarchy, allowing for statistical testing of differences between those low-dimensional manifolds.
